Metric Spaces with Expensive Distances

### 2018-12-05 paper/draft.tex
pgfplots

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{comment}
\usepackage{braket}

@@

We consider two standard operations on metric spaces,
namely the construction of an $\epsilon$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

@@

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm is included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_sparseness}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
    \caption{Results of blind greedy spanner for different dimensions.}
    \label{fig:blind_greedy_only}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
    \caption{Comparison of the four variants of \textsc{BlindRandomBadRatio} algorithm.}
    \label{fig:blind_rbr_variants}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\epsilon$. Data is for 400 normally distributed points in $\R^2$.}
    \label{fig:spanner_eps_dependence}
\end{figure}

@@

The plot \ref{fig:spanner_sparseness}.
shows the number of edges of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
it is important to remember that for all blind spanners the number of edges
is equal to the number of distances that were actually computed, while
for non-blind greedy spanner the latter number is always $n \choose 2$.
The second plot shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

We remark that we obtain an exact nearest neighbor algorithm
when setting $\eps$ to $0$, which means to 
replacing the condition in the if-statement of the loop by
with $a_i\geq v$.

@@

In what follows, we let $c_1,\ldots,c_n$ denote the candidates
at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
With the previous lemma, we can derive an upper bound for the number
of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
as follows.

@@

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
    Next $p_i$ is placed in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$.}
    \label{fig:exact_bad_example}
\end{figure}

>>>>

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{comment}
\usepackage{braket}
\usepackage{pgfplots}

@@

We consider two standard operations on metric spaces,
namely the construction of a $1+\eps$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

@@

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylog time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind quasi-sorted greedy}

    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind quasi-sorted shaker}

    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{WSPD}

    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random}

    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind greedy}

    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Greedy}

\end{axis}
\end{tikzpicture}
\end{centering}
    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_sparseness}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
        legend pos=outer north east]

    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    %\addlegendentry{Blind quasi-sorted greedy}

    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    %\addlegendentry{Blind quasi-sorted shaker}

    %\addplot+ table [x=n_points, y=ratio_wspd, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    %\addlegendentry{WSPD}

    %\addplot+ table [x=n_points, y=ratio_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    %\addlegendentry{Blind random}

    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Blind greedy}

    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Greedy}

\end{axis}
\end{tikzpicture}
\end{centering}
    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Ratio \# edges / \# points for different variants of spanner 
    algorithms. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_ratio}
\end{figure}

\begin{figure}[ht!]
    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
    \caption{Results of blind greedy spanner for different dimensions.}
    \label{fig:blind_greedy_only}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random}

    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, connect first}

    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, connect first, lower bound first}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
    \label{fig:blind_rbr_variants}
\end{figure}

%\begin{figure}[ht]
%    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
%\end{figure}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {eps_dep.csv};
    \addlegendentry{Greedy, dim = 2}

    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {eps_dep.csv};
    \addlegendentry{Blind greedy, dim = 2}

    %\addplot+ table [x=Epsilon, y=edges_greedy_dim_3, col sep=comma] {eps_dep.csv};
    %\addlegendentry{Greedy, dim = 3}

    %\addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_3, col sep=comma] {eps_dep.csv};
    %\addlegendentry{Blind greedy, dim = 3}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\R^2$ and $\R^3$.}
    \label{fig:spanner_eps_dependence}
\end{figure}

%\begin{figure}[ht]
%    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
%\end{figure}

@@

Figure \ref{fig:spanner_sparseness}.
shows the number of edges of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
It is important to remember that for all blind spanners the number of edges
is equal to the number of distances that were actually computed, while
for non-blind greedy spanner the latter number is always $n \choose 2$.
Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

We remark that we obtain an exact nearest neighbor algorithm
when setting $\eps$ to $0$, which means 
replacing the condition in the if-statement of the loop
with $a_i\geq v$.

@@

In what follows, we let $c_i$ denote the candidate
at the end of the $i$-th iteration of the loop, and $v_i$
the distance to $\dist(c_i, q)$, $i = 1, \dots, n$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
With the previous lemma, we can derive an upper bound for the number
of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
as follows.

@@

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
    \label{fig:exact_bad_example}
\end{figure}
